The persuasion of symbolic elements in government public communications, and their moderation by personal relevance

### 2018-07-15 paper3_relevance and symbols_01.Rmd
Add files via upload

A growing public administration literature explores whether government public communications can enhance citizens' trust in public organizations and policies. Whereas most of the priovious studies have focused on the *informative* aspects of government communications [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @grimmelikhuijsen_2015; @ho_2017; @james_2014; @james_2011], recent studies further pointed to the emotive effect of *symbolic* elements entangled in them, such as brand names [@teodoro_2018], logos [@karens_2016], and figures [@alonbarkat_2017]. Such symbols and others exist in almost every government communication, and most prominently in public campaigns, which have been increasingly used by public-sector organizations [@eshuis_2012; @marland_2017].

Persuasion through symbolic communications and branding can be perceived favorably, insofar as it enables governments to engage more effectively with relevant publics, thereby to increase their trust and enhance their collaboration. The effectiveness of symbols, as opposed to communication of "objective" information, stems from their potency to overcome citizens' negative biases. On the other hand, employing a broader perspective of democratic theory, one can argue that in the public-sector domain, the emotive effect of symbols undermines the fundamental normative expectation that citizens' evaluation of their government should be based on their critical judgment. Furthermore, previous studies have demonstrated that the positive effect of symbols can compensate for organizations' poor performance (@alonbarkat_2017) and for logically unpersuasive informative content (@alonbarkat_2018). Accordingly, symbolic communication can create a misalignment between citizens' positive perceptions and the "real world", and mitigate justified public criticism of government failures.

@@

The results of this study, if generalizable, have significant implications. They challenge the optimistic view of government direct communications with citizens as a means to inform citizens about government organizations actions and promote transparency and accountability. Instead, they highlight the significant role played by superficial symbols in public communications in shaping citizens view of governments, as suggested by recent scholars of government branding. I demonstrate that citizens can be persuaded by government branding and symbolic communication even with regard to policy issues that have direct implications on their own lives. These conclusions buttress the concern that government organizations will strategically use symbolic communications to manipulate public opinion, and cause citizens to view them more favorably regardless (or even in contrast) to their actual capacities and achievements. 
How can we explain the finding that the positive effect of the real symbols was stronger among Haifa-Bay residents? Given that this pattern is also evident for the recycling policy, it is unlikely that it is linked to differences in perceived personal relevance. Rather, they can be explained by the lower baseline of Haifa-Bay subjects' trust in both policies, presumably due to their negative prior attitudes towards EPM. Accordingly, it could be the case that symbolic elements are more effective in shifting people from low to medium levels of trust (i.e. as in the case of Haifa-Bay subjects), than shifting them from medium to high levels (i.e. as in the case of center subjects), indicating a ceiling effect. Stated otherwise, symbols may be more effective in reducing citizens' distrust, as opposed to increasing their trust.[^footnote_ceiling_effect] A related explanation is that the symbols distracted the subjects and decreased their inclination to think about the topic. As a result, they caused Haifa-bay subjects to disregard their prior beliefs about EPM and its failure regarding air-pollution.[^footnote_mediation_elaboration]:

@@

>>>>

A growing public administration literature explores whether government public communications can enhance citizens' trust in public organizations and policies. Whereas most of the priovious studies have focused on the *informative* aspects of government communications [e.g. @marvel_2015; @im_2014; @porumbescu_2017; @porumbescu_2016; @grimmelikhuijsen_2015; @ho_2017; @james_2014; @james_2011], recent studies further pointed to the emotive effect of *symbolic* elements entangled in them, such as brand names [@teodoro_2018], logos [@karens_2016], and figures [@alonbarkat_2017]. Such symbols and others exist in almost every government communication, and most prominently in public campaigns and branding activities, which have been increasingly used by public-sector organizations [@eshuis_2012; @marland_2017].

Persuasion through symbolic communications and branding can be perceived favorably, insofar as it enables governments to engage more effectively with relevant publics, thereby to increase their trust and enhance their collaboration. The effectiveness of symbols, as opposed to communication of "objective" information, stems from their potency to overcome citizens' negative biases. On the other hand, employing a broader perspective of democratic theory, one can argue that in the public-sector domain, the emotive effect of symbols undermines the fundamental normative expectation that citizens' evaluation of their government should be based on their critical judgment. Furthermore, previous studies have demonstrated that the positive effect of symbols can compensate for organizations' poor performance [@alonbarkat_2017] and for logically unpersuasive informative content [@alonbarkat_2018]. Accordingly, they can create a misalignment between citizens' positive perceptions and the "real world", and mitigate justified public criticism of government failures.

@@

How can we explain the finding that the positive effect of the real symbols was stronger among Haifa-Bay residents? Given that this pattern is also evident for the recycling policy, it is unlikely that it is linked to differences in perceived personal relevance. Rather, they can be explained by the lower baseline of Haifa-Bay subjects' trust in both policies, presumably due to their negative prior attitudes towards EPM. Accordingly, it could be the case that symbolic elements are more effective in shifting people from low to medium levels of trust (i.e. as in the case of Haifa-Bay subjects), than shifting them from medium to high levels (i.e. as in the case of center subjects), indicating a ceiling effect. Stated otherwise, symbols may be more effective in reducing citizens' distrust, as opposed to increasing their trust.[^footnote_ceiling_effect] A related explanation is that the symbols distracted the subjects and decreased their inclination to think about the topic. As a result, they caused Haifa-bay subjects to disregard their prior beliefs about EPM and its failure regarding air-pollution.[^footnote_mediation_elaboration]:

@@

The results of this study, if generalizable, have significant implications. They challenge the view of government direct communications with citizens as a means to inform citizens about government organizations actions and promote transparency and accountability. Instead, they suggest that more often than not, citizens rely on symbols and other peripheral cues in public communications, even when these communications concern issues that have direct implications on their own lives. Indeed, these findings highlight the role played by symbols in shaping citizens view of governments, in line with the recent literature on government branding. Moreover, they suggest that the boundaries of persuasion via symbols and branding are wider than expected, based on previous social psychology and marketing studies. These conclusions buttress the concern that government organizations will strategically use symbolic communications to manipulate public opinion, and cause citizens to view them more favorably regardless (or even in contrast) to their actual capacities and achievements.
